---
title: "The Classifacation of A Binary Target Variable Using a Binomial Generalized Linear Model"
author: "TJ Smith"
output:
  word_document: default
  html_document: default
---

# Load in needed packages
```{r,include=FALSE}
library(tidyverse)
library(GLMsData)
library(MASS)
library(gridExtra)
library(statmod)
```

#Import Dataset
```{r,include=FALSE}
co <- read.csv("college.csv")
```

# Creating exploratory data plot to obersve the relationship between the variable Outstate vs the target variable
```{r,co-exploratory-plot-1}
pdc <- ggplot(data = co,
            mapping = aes(x = Outstate,
                          y = Target))
pdc <- pdc + geom_jitter(height = 0.02, pch = 1)
pdc <- pdc + geom_smooth(se=FALSE)
pdc <- pdc + labs(x = "Out of State Tuiton Cost",
              y = "Target Variable")
pdc
```

# Getting a summary of the target variable
```{r}
summary(co$Target)
```

# Creating the full model (using all variables)
```{r,include=FALSE}
fullmodel <- glm(Target ~ Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+P.Undergrad+Outstate+Room.Board+Books+Personal+PhD+Terminal+S.F.Ratio+perc.alumni+Expend+Grad.Rate,data=co,family=binomial)
summary(fullmodel)
```

# Using Chisq test to weed out some non-important variables
```{r,include=FALSE}
anova(fullmodel,test="Chisq",data=co)
```

# The next 4 chunks of code include possible models. 
#Used the Wald test to weed out variables to improve model.
```{r,include=FALSE}
adjmodel1 <- glm(Target ~ Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+Outstate+PhD+perc.alumni,data=co,family=binomial)
summary(adjmodel1)
```

```{r,include=FALSE}
adjmodel2 <- glm(Target ~ Apps+Accept+Enroll+Top10perc+Top25perc+F.Undergrad+P.Undergrad+Room.Board+Books+Personal+Terminal+S.F.Ratio+Expend+Grad.Rate,data=co,family=binomial)
summary(adjmodel2)
```

```{r,include=FALSE}
adjmodel3 <- glm(Target ~ Outstate+PhD+perc.alumni+Apps+Enroll+F.Undergrad+Room.Board+Terminal+Expend+Grad.Rate,data=co,family=binomial)
summary(adjmodel3)
```

```{r,include=FALSE}
adjmodel4 <- glm(Target ~ Outstate+PhD+perc.alumni+Apps+F.Undergrad+Expend,data=co,family=binomial)
summary(adjmodel4)
```

# Final model
```{r}
adjmodel5 <- glm(Target ~ Outstate+PhD+perc.alumni+Apps+Expend,data=co,family=binomial)
summary(adjmodel5)
```

The equation for the final model:
$$
logit(mu) = -1.297 + 0.00072*Outstate + 0.0864*perc.alumni + 0.00027*Expend - 0.0835*phd - 0.00053*Apps
$$
# Calculating necesarry columns for model analysis
```{r,include=FALSE}
co$adjmodel5.mu <- predict(adjmodel5,type="response")
co$adjmodel5.eta <- predict(adjmodel5,type="link")
co$adjmodel5.Prediction <- ifelse(co$adjmodel5.mu>0.5,1,0)
co$adjmodel5.rQ1 <- qresid(adjmodel5)
co$adjmodel5.rQ2 <- qresid(adjmodel5)
co$adjmodel5.wr <- co$adjmodel5.eta + resid(adjmodel5,type="working")
```

# Out-of-model plot analysis
```{r}
pom <- ggplot(data = co,
            mapping = aes(x = Room.Board,
                          y = adjmodel5.rQ1))
pom <- pom + geom_point() + geom_smooth(se = TRUE)
pom <- pom + labs(x = "Cost of Room and Board",
              y = "Quantile Residuals",
              title = "Model")

qom <- ggplot(data = co,
            mapping = aes(x = Room.Board,
                          y =adjmodel5.rQ2))
qom <- qom + geom_point() + geom_smooth(se = TRUE)
qom <- qom + labs(x = "Cost of Room and Board",
              y = "Quantile Residuals",
              title = "Model")
grid.arrange(pom, qom, nrow = 1)
```

# Function Family test (Is binomial the right choice?)
```{r}
pt <- ggplot(data = co,
            mapping = aes(sample = adjmodel5.rQ1))
pt <- pt + geom_qq(pch = 1) + geom_qq_line(color = "blue")
pt <- pt + labs(x = "Theoretical Quantiles",
              y = "Sample Quantiles")

qt <- ggplot(data = co,
            mapping = aes(sample = adjmodel5.rQ2))
qt <- qt + geom_qq(pch = 1) + geom_qq_line(color = "blue")
qt <- qt + labs(x = "Theoretical Quantiles",
              y = "Sample Quantiles")
grid.arrange(pt, qt, nrow = 1)
```

# Link function accuracy test
```{r}
p <- ggplot(data = co,
            mapping = aes(x = adjmodel5.eta,
                          y = adjmodel5.wr))
p <- p + geom_point() + geom_smooth(se = TRUE)+ylim(-20,20)
p <- p + labs(x = "Linear Predictor",
              y = "Working Residuals",
              title = "Model")
p
```

# Overall model accuracy predictor
```{r}
p <- ggplot(data = co,
            mapping = aes(x = adjmodel5.mu,
                          y = adjmodel5.rQ1))
p <- p + geom_point() + geom_smooth(se = TRUE)
p <- p + labs(x = "Fitted Values",
              y = "Quantile Residuals",
              title = "Model")

q <- ggplot(data = co,
            mapping = aes(x = adjmodel5.mu,
                          y = adjmodel5.rQ2))
q <- q + geom_point() + geom_smooth(se = TRUE)
q <- q + labs(x = "Fitted Values",
              y = "Quantile Residuals",
              title = "Model")
grid.arrange(p,q, nrow = 1)
```

# Confusion matrix calculation
# Acuraccy, Sensitivity, and Specificity percentages
```{r}
(confmat<- xtabs( ~ Target + adjmodel5.Prediction,
       data = co))
```
  
$$
  \frac{119+346}{119 + 18 + 17 + 346} = \frac{465}{500} = 0.93
$$
$$
  \frac{346}{18 + 346} = \frac{346}{364} = 0.9505
$$

$$
  \frac{119}{119 + 17} = \frac{119}{136} = 0.875
$$
# Function to apply model to new data (Predictive tool)
```{r,co-score-function}
score <- function(dta) {
  Outstate <- co$Outstate
  phd <- co$PhD
  perc.alumni <- co$perc.alumni
  Apps <- co$Apps
  Expend <- co$Expend
  eta <- -1.2978349723 + 0.0007242341*Outstate + 0.0864645073*perc.alumni + 0.0002704450*Expend - 0.0835232929*phd - 0.0005358043*Apps
  mu <- 1/(1+exp(-eta))
  co$Prediction <- ifelse(mu>0.5 ,1,0)
  return(co)
}
```

# Applying function to our dataset
```{r}
tmps <- score(co)
tmps
```
